---
title: "HockeyML"
author: "Kevin Donohue"
date: "2023-10-25"
output:
  word_document: default
  html_document: default
  pdf_document: default
---




This chunk loads in all of my packages:
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}


library(MASS)
library(ggplot2)

library(corrgram)
library(tidyverse)

library(class)
library(caret)

library(rpart)
library(rpart.plot)
```


This chucnk loads in my team dataset
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}

teamdata <- read.csv("C:/Users/kevin/OneDrive/Desktop/PDAT620G/HockeyML/teamdata.csv")
View(teamdata)

```


This chunk removes the asterisk from team names 
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
teamdata$Tm <- gsub("\\*", "", teamdata$Tm)
```



This chunk creates a new Team Abbreviation variable based on the full team names
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
teamdata <- teamdata %>% 
       mutate('Tm' = case_when(Tm == "Boston Bruins" ~ "BOS" ,
  Tm == "Toronto Maple Leafs" ~ "TOR" ,
  Tm == "Tampa Bay Lightning" ~ "TBL" ,
  Tm == "Florida Panthers" ~ "FLA" ,
  Tm == "Buffalo Sabres" ~ "BUF" ,
  Tm == "Ottawa Senators" ~ "OTT" ,
  Tm == "Detroit Red Wings" ~ "DET" ,
  Tm == "Montreal Canadiens" ~ "MTL" ,
  Tm == "Carolina Hurricanes" ~ "CAR" ,
  Tm == "New Jersey Devils" ~ "NJD" ,
  Tm == "New York Rangers" ~ "NYR" ,
  Tm == "New York Islanders" ~ "NYI" ,
  Tm == "Pittsburgh Penguins" ~ "PIT" ,
  Tm == "Washington Capitals" ~ "WSH" ,
  Tm == "Philadelphia Flyers" ~ "PHI" ,
  Tm == "Columbus Blue Jackets" ~ "CBJ" ,
  Tm == "Colorado Avalanche" ~ "COL" ,
  Tm == "Dallas Stars" ~ "DAL" ,
  Tm == "Minnesota Wild" ~ "MIN" ,
  Tm == "Winnipeg Jets" ~ "WPG" ,
  Tm == "Nashville Predators" ~ "NSH" ,
  Tm == "St. Louis Blues" ~ "STL" ,
  Tm == "Arizona Coyotes" ~ "ARI" ,
  Tm == "Phoenix Coyotes" ~ "PHX" ,
  Tm == "Chicago Blackhawks" ~ "CHI" ,
  Tm == "Vegas Golden Knights" ~ "VEG" , 
  Tm == "Edmonton Oilers" ~ "EDM" ,
  Tm == "Los Angeles Kings" ~ "LAK" ,
  Tm == "Seattle Kraken" ~ "SEA" ,
  Tm == "Calgary Flames" ~ "CGY" ,
  Tm == "Vancouver Canucks" ~ "VAN" ,
  Tm == "San Jose Sharks" ~ "SJS" ,
  Tm == "Anaheim Ducks" ~ "ANA" ,
  Tm == "Atlanta Thrashers" ~ "ATL" ))

#View(teamdata)
```




This chunk loads in my goalie data
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
goaliedata <- read.csv("C:/Users/kevin/OneDrive/Desktop/PDAT620G/HockeyML/goaliedata.csv")
head(goaliedata)
```



This chunk filters out the value "TOT" from the team variable as these are duplicates
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
goaliedata <- goaliedata %>%
  filter(!Tm %in% c( "TOT"))
```


This chunk groups the goalie data by team and year, and sums the numerical variables
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
consolidated_goalie <- goaliedata %>%
  group_by(Tm, Yr) %>%
  summarise(across(where(is.numeric), sum))


#View(consolidated_goalie)
```


This chunk loads in my skater data
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep="none"}
skaterdata <- read.csv("C:/Users/kevin/OneDrive/Desktop/PDAT620G/HockeyML/skaterdata.csv")
head(skaterdata)
```

This chunk filters out unneccesary team values
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
skaterdata <- skaterdata %>%
  filter(!Tm %in% c("F", "D", "TOT"))
```

This chunk groups the skater data by team and year, and sums the numerical variables
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
consolidated_skater <- skaterdata %>%
  group_by(Tm, Yr) %>%
  summarise(across(where(is.numeric), sum))


View(consolidated_skater)
```




This chunk joins the three datasets into the full data set and renames select variables
```{r, message=FALSE, warning=FALSE, results = 'hide',fig.keep="none"}
almostfulldata <- full_join(consolidated_skater, consolidated_goalie, by = c("Tm", "Yr"))




fulldata <- full_join(almostfulldata, teamdata, by = c("Tm", "Yr"))
view(fulldata)


fulldata <- select(fulldata, GF, GA.y, SOS, OL, W.y, SA, SV, X..., PIM.x, S, HIT, FOW, Playoffs) %>%
  rename(
  GA = GA.y, W = W.y, PM = X..., PIM = PIM.x)

#View(fulldata)

#write_csv(fulldata, "fulldata.csv")


```

POST DATA CLEANING:


This chunk loads in the full dataset and creates a corr plot of the variables
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep="none"}
fulldata <- read.csv("C:/Users/kevin/OneDrive/Desktop/PDAT620G/HockeyML/fulldata.csv")
head(fulldata)
corrgram(fulldata, lower.panel=panel.shade, upper.panel=panel.cor, main = "")
```


This dataset identifies which variables have NA values and how many
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}




na_counts <- count_NA(fulldata)


print(na_counts)

```


This chunk shows the imputation of the median for th HIT and FOW variables
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}


fulldata$HIT[is.na(fulldata$HIT)] <- median(fulldata$HIT, na.rm = TRUE)
fulldata$FOW[is.na(fulldata$FOW)] <- median(fulldata$FOW, na.rm = TRUE)


```



This chunk creates testing and training sets 

```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
set.seed(123)
splitIndex <- createDataPartition(fulldata$W, p = 0.8, list = FALSE)
train_data <- fulldata[splitIndex, ]
test_data <- fulldata[-splitIndex, ]
```

This chunk creates my model and checks for VIF values
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
model <- lm(W ~ GF  +  SV + PIM + S + HIT + FOW + OL, data = train_data[, sapply(train_data, is.numeric)])
vif_values <- vif(model)


print(vif_values)

```


ASSUMPTIONS:

This chunk makes the Residuals vs. Fitted Values Plot (or Residuals vs. Predicted Values).
its purpose is to check for LINEARITY and INDEPENDECE
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
plot(model, which = 1)
```

This chunk creates the Normal Q-Q Plot:
its purpose is to check the NORMALITY of residuals.

```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
qqnorm(residuals(model))
qqline(residuals(model))
```




The chunk creates the Scale-Location (or Spread-Location) Plot:
its purpose is to check for homoscedasticity (constant variance of residuals).

```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
sqrt_abs_standardized_residuals <- sqrt(abs(rstandard(model)))
plot(fitted(model), sqrt_abs_standardized_residuals, ylab = "sqrt(|Standardized Residuals|)", xlab = "Fitted Values", main = "Scale-Location Plot")
abline(h = mean(sqrt_abs_standardized_residuals))
```




This chunk is for scaling/normalizing data:
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}
train_data[, sapply(train_data, is.numeric)] <- scale(train_data[, sapply(train_data, is.numeric)])
test_data[, sapply(test_data, is.numeric)] <- scale(test_data[, sapply(test_data, is.numeric)])
  
```



This chunk evaluates the regression model by making predictions and calculating error and R-squared
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}


model <- lm(W ~ GF  +  SV + PIM + S + HIT + FOW + OL, data = train_data)


summary(model)


predictions <- predict(model, newdata = test_data[, sapply(test_data, is.numeric)])


mae <- mean(abs(predictions - test_data$W))
mse <- mean((predictions - test_data$W)^2)
rmse <- sqrt(mse)



cat("Mean Absolute Error:", mae, "\n")
cat("Mean Squared Error:", mse, "\n")
cat("Root Mean Squared Error:", rmse, "\n")
#cat("R-squared:", rsquared, "\n")
summary(model)
```



his chunk evaluates the CART model by making predictions and calculating error and R-squared
```{r, message=FALSE, warning=FALSE, results = 'hide', fig.keep = "none"}

model2 <- rpart(W ~ GF  +  SV + PIM + S + HIT + FOW + OL, data = train_data)

rpart.plot(model2)


predictions2 <- predict(model2, newdata = test_data)


mae2 <- mean(abs(predictions2 - test_data$W))
mse2 <- mean((predictions2 - test_data$W)^2)
rmse2 <- sqrt(mse2)
rsquared2 <- cor(predictions2, test_data$W)^2  # R-squared for random forest

cat("Mean Absolute Error:", mae2, "\n")
cat("Mean Squared Error:", mse2, "\n")
cat("Root Mean Squared Error:", rmse2, "\n")
cat("R-squared:", rsquared2, "\n")


```


